Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2024-12-12 manuscript/paper.tex

\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG
visualisation, genome browser} \maketitle \section{Introduction} Visualisation
is an essential component of data analysis pipelines. In the past few decades,
genome browsers have become synonymous with sequence visualisation, by
providing user-friendly, interactive interfaces to navigate genomic sequences
and annotations. From early standalone tools to manage \textit{C. elegans} data
\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
with the changing landscape of genomics. Modern genome browsers can load
datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
The data security and scalability offered by this architecture ensures that
genome browsers continue to be relevant in the biobank era.

Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
are related to each other at each position of the genome in a recombining
species~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}, and there has been significant recent progress 
on inference through a range of approaches~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,zhang2023biobank,gunnarsson2024scalable,
deng2024robust}.

Biobank-driven research has accelerated the development of several
technologies, including the ability to infer Ancestral Recombination Graphs
(ARGs) from large-scale genetic data
% \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
ARGs encode the interwoven paths of genetic inheritance resulting from
recombination, and can be represented as genetic genealogies along the sequence
\citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
ancestral processes influencing the samples, and have garnered great interest
in several applications in population and statistical genetics. These include
the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
inference is increasingly becoming a part of genetic data analysis pipelines.
Yet, ARG visualisation, which is crucial for downstream analyses remains
challenging due to the complexity of inferences from empirical data. Existing
visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
semantic representation of ARGs, but such representations have limited utility
in informing downstream analyses.

>>>>

\keywords{Ancestral Recombination Graphs, Holoviz, biobank-scale ARG
visualisation, genome browser} \maketitle

\section{Introduction} 
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a recombining
species, and are currently the subject of intense research 
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
gunnarsson2024scalable,deng2024robust} that practical application has 
become possible. Applications to varied questions such as the spatial 
location of genetic
ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},

as well as the use of ARGs as training data for machine learning 
methods~\citep{whitehouse2024tree} depend critically on the \emph{accuracy}
of the inferred ARGs. Yet, despite this explosion of downstream methods using
inferred ARGs, there is a critical lack of practical tools to help
researchers make informed assesments about the quality of their inferred ARGs.

% Varied applications to
% simulation~\citep{kelleher2016efficient,baumdicker2021efficient,Anderson-Trocme2023}

Visualisation
is an essential component of data analysis pipelines. In the past few decades,
genome browsers have become synonymous with sequence visualisation, by
providing user-friendly, interactive interfaces to navigate genomic sequences
and annotations. From early standalone tools to manage \textit{C. elegans} data
\citep{Eeckman1995}, to modern web-apps for vertebrate genomes
\citep{Birney2004,Rangwala2024,Nassar2023}, browsers have continued to evolve
with the changing landscape of genomics. Modern genome browsers can load
datasets from a URL or local file system \citep{IGV-web,Lee2013,Robinson2023}.
The data security and scalability offered by this architecture ensures that
genome browsers continue to be relevant in the biobank era.

% Biobank-driven research has accelerated the development of several
% technologies, including the ability to infer Ancestral Recombination Graphs
% (ARGs) from large-scale genetic data
% % \citep{Kelleher2019,Speidel2019,Zhang2023,Gunnarsson2024.08.31.610248,Deng2024.03.16.585351}.
% ARGs encode the interwoven paths of genetic inheritance resulting from
% recombination, and can be represented as genetic genealogies along the sequence
% \citep{wong2024general,Hudson1983,Griffiths1997}. They capture rich detail about
% ancestral processes influencing the samples, and have garnered great interest
% in several applications in population and statistical genetics. These include
% the detection of ultra-rare genetic variation \citep{zhang2023biobank}, the mapping of
% quantitative trait loci \citep{Link2023}, and the sparse and efficient modeling
% of linkage disequilibrium \citep{Nowbandegani2023}. It is evident that ARG
% inference is increasingly becoming a part of genetic data analysis pipelines.
% Yet, ARG visualisation, which is crucial for downstream analyses remains
% challenging due to the complexity of inferences from empirical data. Existing
% visualisation tools (see \hyperref[sec:Discussion]{Discussion}) focus on the
% semantic representation of ARGs, but such representations have limited utility
% in informing downstream analyses.
